Translation of B Implementations to the LLVM: Specification

### 2014-03-18 DOCUMENTATION/specification.tex
Completed a first draft for full specification of initialization.

%TODO It is deprecated 
%[[[TODO: DEPRECATED ]]]

Initialization is done is two steps: first all imported modules are
initialized,  second the initialization clause is executed.
\begin{align*}
  \trad{i}{Init} \isdef
  & \llvm{"define void @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
  & \quad\llvm{\PH{\trad{i, i.impl}{\ListOf{InitImp}}}} \\
  & \quad\llvm{\PH{\trad{i.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
  & \quad\llvm{exit: ret void \nl} \\
  & \llvm{\}"}
\end{align*}
The initialization of an imported module is performed by a call to the
corresponding initialization function:
\begin{align*}
  \lefteqn{\trad{\B{impl}, \B{impo}}{InitImp} \isdef} \\
  & \LET v = \variable \AND \B{m} = \B{impo.mach} \IN \\
  & \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{impl.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{impo})}\nl}\\
  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m.impl})}*
    \PH{\ell}) \nl"}
\end{align*}

>>>>

\newcommand{\argname}[0]{\ensuremath{\mathsf{\nabla}}}
\newcommand{\components}[0]{\ensuremath{\textsf{st-comp*}}}

Initialization of a module is implemented by a LLVM function that takes as
parameters: the address of the aggregate representing the state of the module,
which we name \llvm{\%self}, and the addressses of the aggregates representing
the state of each component of the module. 
We first define a naming function $\argname$ having as domain the list of
\emph{stateful} module components:
\begin{align*}
  \argname(l) \isdef
  & \lambda c \cdot \llvm{"\%arg\PH{\mathit{index}(c, l)}"}
\end{align*}
So \llvm{\%arg0}, \llvm{\%arg1}, \ldots are the identifiers for the parameters
representing the components of the initialized module.

More precisely, the specification needs to make a distinction between stateful
and stateless components, as only the former are represented by a LLVM
aggregate. We use the expression $\components(\B{n}) \isdef
\mathit{remove}(\trimportedinstances(\B{n}), \stateless)$ to denote the list of
stateful components of $\B{n}$, so that $\argname(\components(\B{n}))$ is the
function that, given a stateful component $c$ of \B{n}, returns the parameter of
the initialization function of $c$. We are now ready to specify the 
signature of this initialization function, using an auxiliary translation
function to produce the LLVM parameter type and name for each stateful
component $\B{c}$:
\begin{align*}
\trad{\B{n}}{init-sig} \isdef &
\llvm{"\PH{\init(\B{n})}(\PH{\stateref(\B{n})} \%self, 
  \PH{\trad{\components(\B{n}), \argname(\components(\B{n}))}{\ListOf{arg}}})"} \\
\trad{\B{c}, \argname}{arg} \isdef & \llvm{"\stateref(\B{c.root}) \argname(\B{c})"}
\end{align*}

The body of this function is composed as follows: firstly, frame stack memory is
allocated, to store the representation of the local variables in the
initialization; second, all the directly imported, stateful, component modules
are bound to the main module's elements; third, the directly imported components
are initialized, by calling the corresponding initialization function; last, the
instructions composing the initialization clause are executed:

\begin{align*}
  \trad{\B{n}}{\B{init}}_{\text{interface}} \isdef
  & \llvm{"define void \PH{\trad{\B{n}}{init-sig}} \{ \nl"} \\
  & \quad\llvm{entry: \nl} \\
  & \quad\llvm{\PH{\trad{n.init}{alloc}}} \\
  & \quad\llvm{\PH{\trad{n, \importedinstances(n), \argname(\components(\B{n}))}{\ListOf{BindImp}}}} \\
  & \quad\llvm{\PH{\trad{\importedinstances(n), \argname(\components(\B{n}))}{\ListOf{InitImp}}}} \\
  & \quad\llvm{\PH{\trad{n.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
  & \quad\llvm{exit: ret void \nl} \\
  & \llvm{\}"}
\end{align*}

The translation for the memory allocation and the instructions are specified in
section~\ref{sec:trad-instr}.

The aggregate representing the state of the root module has one element for
each directly imported stateful module. The initialization function is
responsible for binding this element with the corresponding argument.
\begin{align*}
  \lefteqn{\trad{\B{n}, \B{c}, \argname}{BindImp} \isdef} \\
  & \LET v = \variable \IN \\
  & \quad \llvm{"\PH{v} = getelementptr \PH{\stateref(\B{n})} \%self\$, i32 0, i32 \PH{\idx(\B{c})}\nl}\\
  & \quad \llvm{store \PH{\stateref(\B{c.root})} \PH{\argname(\B{c})}, 
    \PH{\stateref(\B{c.root})}* \PH{v}\nl"}
\end{align*}

In general, the initialization function of a component module $c$ has as
argument the address of the aggregate representing the module, as well as the
addresses of the aggregates representing the modules in the importation tree of
$c$.
\begin{align*}
  \trad{\B{c}, \argname}{InitImp} \isdef&
  \llvm{call void \PH{\init(\B{c})}(\PH{\trad{c}{arg}}, \trad{\components(\B{c}), \argname}{\ListOf{arg}}) \nl"} \\
\trad{\B{c}, \argname}{arg} \isdef& 
  \llvm{"\PH{\stateref(\B{c.root})} \PH{\argname(\B{c})}"}
\end{align*}
